---
title: "iNat_API"
execute-dir: project
format: 
  html:
    toc: true
    toc-location: right
    smooth-scroll: true
    html-math-method: katex
    code-fold: true
editor: source
author: 'Florencia Grattarola'
date: 2022-12-12
---

## First steps with the iNaturalist API

I was interested to know how many of the users generating records on [NaturalistaUY]( https://www.naturalista.uy) (the Uruguayan site of iNaturalist), are, in fact, Uruguayan users. With **Rodrigo Montiel**, we are assessing the profile of observers in Uruguay (check his repo [here]( https://github.com/Rodrigo-Montiel/TesisNaturalistaUY)). So, we need to detect those users that are not from Uruguay and remove the data generated by them from our dataset. This was a perfect use case to test the [iNat API]( https://api.inaturalist.org/v1/docs/), which, so far, I hadn’t explored.

Basically, this API enables you to query the **iNaturalist** database by using different methods, parameters and values. For instance, you can **search** and **fetch** data on observers, such as number of observations (`observation_count`) or number of species observed (`species_count`), by providing a username (`user_login`). The result is a JSON that you can parse and analyse.

### Let’s get on with it!

To start with an example, I'm going to use my own username (`flo_grattarola`) as value for the `user_login` parameter to do a query, using `GET /observations/observers` ([see it in the API](https://api.inaturalist.org/v1/docs/#!/Observations/get_observations_observers)), which '*returns observers of observations matching the search criteria and the count of observations and distinct taxa of rank species they have observed*'. 

So, by running the following query:
`'https://api.inaturalist.org/v1/observations/observers?user_login=flo_grattarola'`

We get as output, the following data in JSON format:

```{r}
#| echo: false
#| message: false
#| label: json-example
#| warning: false

raw_json <- '
{
  "total_results": 1,
  "page": 1,
  "per_page": 500,
  "results": [
    {
      "user_id": 736016,
      "observation_count": 3649,
      "species_count": 1246,
      "user": {
        "id": 736016,
        "login": "flo_grattarola",
        "spam": false,
        "suspended": false,
        "created_at": "2017-12-15T15:54:34+00:00",
        "login_autocomplete": "flo_grattarola",
        "login_exact": "flo_grattarola",
        "name": "Florencia Grattarola",
        "name_autocomplete": "Florencia Grattarola",
        "orcid": "https://orcid.org/0000-0001-8282-5732",
        "icon": "https://static.inaturalist.org/attachments/users/icons/736016/thumb.jpeg?1513353273",
        "observations_count": 3649,
        "identifications_count": 5779,
        "journal_posts_count": 1,
        "activity_count": 9429,
        "species_count": 1419,
        "universal_search_rank": 3649,
        "roles": [
          "curator"
        ],
        "site_id": 28,
        "icon_url": "https://static.inaturalist.org/attachments/users/icons/736016/medium.jpeg?1513353273"
      }
    }
  ]
}'
listviewer::jsonedit(raw_json, height = "700px", mode = "code")
```
<br>
Great! This query gives us all the data we need, especially the `observation_count` value for each user. 

### Assessing users in NaturalistaUY

```{r}
#| echo: false
#| message: false
#| label: data
#| warning: false

library(knitr)
library(tidyverse)

NatUY_users <- readRDS('../../datos/usuarios_dataset.rds') %>% 
  select(user_id, user_login, observation_count_UY=registros) 
```

First, we need to download the data from [naturalista.uy](naturalista.uy) (ours was downloaded on 2022-10-27). Then, we calculate the number of observations made by each user in Uruguay (`observation_count_UY`), and keep the `user_id` and `user_login` variables. You can do this by using functions `group_by()` and `count()`. You could also do this using the API, but in our case, we already had the data.  

We found a total of **`r nrow(NatUY_users)`** users in NaturalistaUY. The user with largest number of records has **`r max(NatUY_users$observation_count_UY)`** observations and on average users have uploaded **`r round(mean(NatUY_users$observation_count_UY),1)`** records to iNat. 

Here's a glance of the data:

```{r}
#| echo: false
#| message: false
#| label: data-table
#| warning: false
NatUY_users_selection <- NatUY_users %>% 
  filter(user_id %in% c(1255162, 3262606, 2988066, 2640130, 11503, 
                        4109374, 1469479, 1368063, 1569449, 2348924)) 

NatUY_users_selection %>% kable()
```


### Create a function to run the query for multiple users

The idea now is to be able to run the test query from above for **all** the users of NaturalistaUY, get their `observation_count`s and compare them with the number of observations these users have in Uruguay (i.e., proportion of observations recorded in Uruguay vs in the rest of the world).   

The function `get_observers_num_observations()` takes a list of users' login names (`user_login`) and returns a tibble with `user_id`, `user_login` and `observation_count_iNat`. This last count is the total number of observations of the users in the platform.  

An **important consideration** to using this API is that we could flood it by querying all users together, as iNat limits the API usage to a max of 100 requests per minute. So, we need to create a delay in the fetching process. We will do this by pausing the query for 10 seconds every ten rows, using `Sys.sleep()`.  

Here's the function:  

```{r}
#| label: function
#| echo: true
#| eval: true
#| code-fold: false
#| warning: false
#| message: false


library(httr)
library(jsonlite)
library(tidyverse)

get_observers_num_observations <- function(user_login_list){
  observers_num_observations <- tibble(user_id = numeric(), 
                                       user_login = character(),
                                       observation_count_iNat = numeric())
 
  num_results <- 1  # se usa para dormir la llamada a la API y para imprimir en consola el progreso
  
  for (user_login in user_login_list) {
    
    if ((num_results %% 10) + 10 == 10) { 
      Sys.sleep(10) # La API necesita un delay porque si no da error. Cada 10 users, el código para 10 segundos
    }
    
    call <- paste0("https://api.inaturalist.org/v1/observations/observers?user_login=", user_login)
    
    get_json_call <- GET(url = call) %>%
      content(as = "text") %>% fromJSON(flatten = TRUE)
    
    if (is.null(get_json_call)) {
      observer_num_observations <- tibble(user_id = NA, 
                                          user_login = user_login,
                                          observation_count_iNat = NA)
      observers_num_observations <- rbind(observers_num_observations, observer_num_observations)
      cat(num_results, 'usuario:', user_login, '--> NOT FOUND', '\n')
    }
    else {
      results <- as_tibble(get_json_call$results) 
      observer_num_observations <- tibble(user_id = results$user_id,
                                          user_login = results$user.login,
                                          observation_count_iNat = results$observation_count)

      observers_num_observations <- rbind(observers_num_observations, observer_num_observations)
      cat(num_results, 'usuario:', user_login, '--> DONE', '\n')
    }
    num_results <- nrow(observers_num_observations) + 1 
  }
  return(observers_num_observations)
}
```

It is probably written in a too complicated way, but it does the job 🥹 

### Let's run it

To run the function, we need to provide a list of users' `user_login`s. Let's use as an example the previous list.

```{r}
#| echo: false
#| message: false
#| label: users-login
#| warning: false
NatUY_users_selection$user_login
```

When we run it, the function prints in the console the users' `user_login` that it is assessing, so we can have an idea of the progress. If the user is found it will print the name and `--> DONE `, while if it's not found, it will return `--> NOT FOUND`.  

```{r}
#| echo: true
#| message: false
#| label: run
#| warning: false
NatUY_users_assessment <- get_observers_num_observations(NatUY_users_selection$user_login)
```

### How do results look?

In the end we get a table with the counts.

```{r}
#| echo: false
#| message: false
#| label: results
#| warning: false
NatUY_users_assessment %>% kable
```

Finally, we merge the results with our original table and count the proportion of records from each user that were recorded in Uruguay. We can even make a guess of who is Uruguayan (those that did more than %30 of their observations in Uruguay), see variable `esUruguaye`.

```{r}
#| echo: true
#| message: false
#| label: merge
#| warning: false

left_join(NatUY_users_selection, NatUY_users_assessment) %>% 
  mutate(proporcion = round(observation_count_UY*100/observation_count_iNat, 3),
         esUruguaye = ifelse(proporcion>=33, 'si', 'no')) %>% kable
```

### Conclusions

```{r}
#| echo: false
#| message: false
#| label: conclusions
#| warning: false

users <- read_csv('../../datos/NatUY_users_UY.csv') %>% 
  mutate(proporcion = round(registros*100/observation_count, 3),
         esUruguaye = ifelse(proporcion>=33, 'si', 'no')) %>%
  select(user_id, user_login, observation_count_iNat=observation_count, 
         observation_count_UY=registros, proporcion, esUruguaye)
```

From the total of **`r nrow(users)`** users, **`r nrow(users[users$esUruguaye=='si',])`** are Uruguayans (i.e., have recorded more than 1/3 of their observations in Uruguay), while **`r nrow(users[users$esUruguaye=='no',])`** are not. We also found **`r nrow(users[is.na(users$esUruguaye),])`** users that seem to have deleted their accounts in the platform and, thus, they were not found.

#### That's all folks! 
